---
title: "Biostatistical Method HW5"
author: "Chenxi Liu"
date: "2026-03-29"
output: html_document
---
# Question 1
```{r}
# Load the data
crab <- read.table("HW5-crab.txt", header = TRUE)

head(crab)
dim(crab)
summary(crab)
```
### (a) Fit a Poisson model (M1) with log link with W as the single predictor. Check goodness of fit and interpret the model.
```{r}
M1 <- glm(Sa ~ W, family = poisson(link = "log"), data = crab)
summary(M1)
coef(M1)
exp(coef(M1))

deviance(M1)
df.residual(M1)

1 - pchisq(deviance(M1), df.residual(M1))
pearson_M1 <- sum(residuals(M1, type = "pearson")^2)
pearson_M1
1 - pchisq(pearson_M1, df.residual(M1))
```
(a) I fitted the Poisson regression model Sa ~ W with a log link. The fitted model is $\log(\mu_i) = -3.3048 + 0.1640W_i$, where $\mu_i = E(Sa_i \mid W_i)$. Equivalently, $\mu_i = \exp(-3.3048 + 0.1640W_i)$. Since $\exp(0.1640) = 1.178$, for each 1-unit increase in carapace width, the expected number of satellites is multiplied by 1.178, corresponding to about a 17.8% increase in the expected number of satellites. The effect of width is statistically significant. 
However, the goodness-of-fit is poor because both the residual deviance and Pearson chi-square statistic are much larger than their degrees of freedom,p = 0,indicating lack of fit. Therefore, although width is positively associated with the number of satellites, the Poisson model with width alone does not fit the data well.

### (b) Poisson model with width and weight and compared with model in (a)
```{r}
M2 = glm(Sa ~ W + Wt, family = poisson(link = "log"), data = crab)
coef(M2)
exp(coef(M2))
# Compare M2 and M1
anova(M1, M2, test = "Chisq")
```
The fitted Poisson regression model is log(E(Sa | W, Wt)) = -1.2917 + 0.0459W + 0.4474Wt. Holding weight fixed, a 1-unit increase in width multiplies the expected number of satellites by exp(0.0459) = 1.047, corresponding to a 4.7% increase. Holding width fixed, a 1-unit increase in weight multiplies the expected number of satellites by exp(0.4474) = 1.564, corresponding to a 56.4% increase. The likelihood ratio test comparing M2 to M1 gives a deviance difference of 7.9934 with 1 degree of freedom and p = 0.0047, so M2 provides a significantly better fit than M1.

### (c) Check over dispersion in M2.
```{r}
# (c) Check overdispersion in M2

# Pearson chi-square statistic
X2 <- sum(residuals(M2, type = "pearson")^2)
X2
df <- df.residual(M2)
df
phi <- X2 / df
phi
1 - pchisq(X2, df)

deviance(M2) / df.residual(M2)

M2.q <- glm(Sa ~ W + Wt, family = quasipoisson(link = "log"), data = crab)
summary(M2.q)
coef(M2.q)
confint.default(M2.q)

exp(coef(M2.q))
exp(confint.default(M2.q))
```
The Pearson chi-square statistic gives phi = 3.156, and the deviance-based estimate is 559.89 / 170 = 3.293, both of which are much greater than 1. This indicates substantial overdispersion, so the Poisson model is not adequate. 

After adjusting for overdispersion using a quasi-Poisson model, the fitted mean model remains log(E(Sa | W, Wt)) = -1.2917 + 0.0459W + 0.4474Wt. Thus, holding weight fixed, a 1-unit increase in width multiplies the expected number of satellites by exp(0.0459) = 1.047, and holding width fixed, a 1-unit increase in weight multiplies the expected number of satellites by exp(0.4474) = 1.564. 

However, after adjustment the standard errors are larger, and the 95% confidence intervals for both effects include 1, with width given by exp((-0.1170, 0.2088)) = (0.890, 1.232) and weight given by exp((-0.1050, 0.9998)) = (0.900, 2.718). Therefore, after accounting for overdispersion, neither width nor weight is statistically significant.

# Problem 2
```{r}
#Data import
dat = read.table("HW5-parasite.txt", header = TRUE)
dat = dat[, c("Intensity", "Area", "Year", "Length")]
dat = na.omit(dat)

dat$Area = factor(dat$Area)
dat$Year = factor(dat$Year)
```
### (a) Poisson model
```{r}
fit.pois = glm(Intensity ~ Area + Year + Length,
               family = poisson(link = "log"), data = dat)
summary(fit.pois)
exp(coef(fit.pois))
```
The fitted model is `log(E(Intensity | Area, Year, Length)) = 2.6432 - 0.2120 I(Area=2) - 0.1169 I(Area=3) + 1.4049 I(Area=4) + 0.6703 I(Year=2000) - 0.2181 I(Year=2001) - 0.0284 Length`, with Area 1 and Year 1999 as the reference categories. Controlling for year and length, fish in Area 2, Area 3, and Area 4 have expected parasite intensity that is 0.809, 0.890, and 4.075 times that of fish in Area 1, respectively. 

Controlling for area and length, fish in Year 2000 and Year 2001 have expected parasite intensity that is 1.955 and 0.804 times that of fish in Year 1999, respectively. Controlling for area and year, each 1-unit increase in length is associated with expected parasite intensity being multiplied by 0.972.

### Problem (b) Goodness of fit
```{r}
deviance(fit.pois)
df.residual(fit.pois)
1 - pchisq(deviance(fit.pois), df.residual(fit.pois))

pearson = sum(residuals(fit.pois, type = "pearson")^2)
1 - pchisq(pearson, df.residual(fit.pois))
```
The goodness-of-fit of the Poisson model is poor. The residual deviance is 19152.8 with 1184 degrees of freedom, and the corresponding deviance goodness-of-fit test gives p ≈ 0. The Pearson chi-square goodness-of-fit test also gives p ≈ 0. Therefore, we reject the null hypothesis that the Poisson model fits the data adequately. The model shows clear lack of fit, suggesting that the data may exhibit overdispersion or excess zeros, so a zero-inflated model is reasonable to consider in part (c).

### (c) Zero-inflated Poisson model
```{r}
library(pscl)
fit.zip = zeroinfl(Intensity ~ Area + Year + Length |
                   Area + Year + Length,
                   dist = "poisson", data = dat)
summary(fit.zip)
exp(coef(fit.zip))
```
In the count part of the zero-inflated Poisson model, controlling for year and length, fish in Area 2, Area 3, and Area 4 have expected parasite intensity that is 1.308, 1.158, and 2.572 times that of fish in Area 1, respectively. Controlling for area and length, fish in Year 2000 and Year 2001 have expected parasite intensity that is 1.480 and 0.956 times that of fish in Year 1999, respectively. Controlling for area and year, a 1-unit increase in length multiplies the expected parasite intensity by 0.964.

In the zero-inflation part, controlling for year and length, fish in Area 2 and Area 3 have odds of belonging to the excess-zero group that are 2.052 and 1.930 times those of fish in Area 1, while fish in Area 4 have odds that are 0.360 times those of fish in Area 1. Controlling for area and length, fish in Year 2000 and Year 2001 have odds of belonging to the excess-zero group that are 0.471 and 1.579 times those of fish in Year 1999, respectively. Controlling for area and year, a 1-unit increase in length multiplies these odds by 0.990.