---
title: "Biostatistical Method HW5"
author: "Chenxi Liu"
date: "2026-03-29"
output: html_document
---
# Question 1
```{r}
# Load the data
crab <- read.table("HW5-crab.txt", header = TRUE)

head(crab)
dim(crab)
summary(crab)
```
### (a) Fit a Poisson model (M1) with log link with W as the single predictor. Check goodness of fit and interpret the model.
```{r}
M1 <- glm(Sa ~ W, family = poisson(link = "log"), data = crab)
summary(M1)
coef(M1)
exp(coef(M1))

deviance(M1)
df.residual(M1)

1 - pchisq(deviance(M1), df.residual(M1))
pearson_M1 <- sum(residuals(M1, type = "pearson")^2)
pearson_M1
1 - pchisq(pearson_M1, df.residual(M1))
```
(a) I fitted the Poisson regression model Sa ~ W with a log link. The fitted model is $\log(\mu_i) = -3.3048 + 0.1640W_i$, where $\mu_i = E(Sa_i \mid W_i)$. Equivalently, $\mu_i = \exp(-3.3048 + 0.1640W_i)$. Since $\exp(0.1640) = 1.178$, for each 1-unit increase in carapace width, the expected number of satellites is multiplied by 1.178, corresponding to about a 17.8% increase in the expected number of satellites. The effect of width is statistically significant. 
However, the goodness-of-fit is poor because both the residual deviance and Pearson chi-square statistic are much larger than their degrees of freedom,p = 0,indicating lack of fit. Therefore, although width is positively associated with the number of satellites, the Poisson model with width alone does not fit the data well.

### (b) Poisson model with width and weight and compared with model in (a)
```{r}
M2 = glm(Sa ~ W + Wt, family = poisson(link = "log"), data = crab)
coef(M2)
exp(coef(M2))
# Compare M2 and M1
anova(M1, M2, test = "Chisq")
```
The fitted Poisson regression model is log(E(Sa | W, Wt)) = -1.2917 + 0.0459W + 0.4474Wt. Holding weight fixed, a 1-unit increase in width multiplies the expected number of satellites by exp(0.0459) = 1.047, corresponding to a 4.7% increase. Holding width fixed, a 1-unit increase in weight multiplies the expected number of satellites by exp(0.4474) = 1.564, corresponding to a 56.4% increase. The likelihood ratio test comparing M2 to M1 gives a deviance difference of 7.9934 with 1 degree of freedom and p = 0.0047, so M2 provides a significantly better fit than M1.

### (c) Check over dispersion in M2.
```{r}
# (c) Check overdispersion in M2

# Pearson chi-square statistic
X2 <- sum(residuals(M2, type = "pearson")^2)
X2
df <- df.residual(M2)
df
phi <- X2 / df
phi
1 - pchisq(X2, df)

deviance(M2) / df.residual(M2)

M2.q <- glm(Sa ~ W + Wt, family = quasipoisson(link = "log"), data = crab)
summary(M2.q)
coef(M2.q)
confint.default(M2.q)

exp(coef(M2.q))
exp(confint.default(M2.q))
```
The Pearson chi-square statistic gives phi = 3.156, and the deviance-based estimate is 559.89 / 170 = 3.293, both of which are much greater than 1. This indicates substantial overdispersion, so the Poisson model is not adequate. 

After adjusting for overdispersion using a quasi-Poisson model, the fitted mean model remains log(E(Sa | W, Wt)) = -1.2917 + 0.0459W + 0.4474Wt. Thus, holding weight fixed, a 1-unit increase in width multiplies the expected number of satellites by exp(0.0459) = 1.047, and holding width fixed, a 1-unit increase in weight multiplies the expected number of satellites by exp(0.4474) = 1.564. 

However, after adjustment the standard errors are larger, and the 95% confidence intervals for both effects include 1, with width given by exp((-0.1170, 0.2088)) = (0.890, 1.232) and weight given by exp((-0.1050, 0.9998)) = (0.900, 2.718). Therefore, after accounting for overdispersion, neither width nor weight is statistically significant.