---
title: "Biostatistical Method HW5"
author: "Chenxi Liu"
date: "2026-03-29"
output: html_document
---
# Question 1
```{r}
# Load the data
crab <- read.table("HW5-crab.txt", header = TRUE)

head(crab)
dim(crab)
summary(crab)
```
### (a) Fit a Poisson model (M1) with log link with W as the single predictor. Check goodness of fit and interpret the model.
```{r}
M1 <- glm(Sa ~ W, family = poisson(link = "log"), data = crab)
summary(M1)
coef(M1)
exp(coef(M1))

deviance(M1)
df.residual(M1)

1 - pchisq(deviance(M1), df.residual(M1))
pearson_M1 <- sum(residuals(M1, type = "pearson")^2)
pearson_M1
1 - pchisq(pearson_M1, df.residual(M1))
```
(a) I fitted the Poisson regression model Sa ~ W with a log link. The fitted model is $\log(\mu_i) = -3.3048 + 0.1640W_i$, where $\mu_i = E(Sa_i \mid W_i)$. Equivalently, $\mu_i = \exp(-3.3048 + 0.1640W_i)$. Since $\exp(0.1640) = 1.178$, for each 1-unit increase in carapace width, the expected number of satellites is multiplied by 1.178, corresponding to about a 17.8% increase in the expected number of satellites. The effect of width is statistically significant. 
However, the goodness-of-fit is poor because both the residual deviance and Pearson chi-square statistic are much larger than their degrees of freedom,p = 0,indicating lack of fit. Therefore, although width is positively associated with the number of satellites, the Poisson model with width alone does not fit the data well.



